---
title: "Overview of Earthquake Plots"
author: "L. Mitchell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of Earthquake Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Timeline Plots

The `earthquake` package contains functions for creating timeline plots 
and maps of earthquakes. The `eq_timeline` function produces a timeline 
plot built on the `ggplot2` library. An example plot is shown below. 

```{r, fig.height=4, fig.width=8}
eqDataClean <- eq_clean_data("NOAA_earthquakes.txt")
eqDataShort <- eqDataClean %>%
  dplyr::filter(Country %in% c("Mexico","New Zealand") & YEAR >= 2006 & 
                YEAR <= 2010)
eq_timeline(dataset=eqDataShort)
```

Date is represented on the x-axis, Country is represented on the y-axis, 
point size reflects the magnitude (based on the field `EQ_PRIMARY`), and 
fill color reflects the total number of deaths (based on the field 
`TOTAL_DEATHS`). A custom theme with a white background, `eqTheme`, is 
automatically applied. 

If the optional argument `label` is set to `TRUE`, the value of 
`LocalLocation` is added to each point as a text label:

```{r, fig.height=5, fig.width=8}
eq_timeline(dataset=eqDataShort, label=TRUE)
```

The optional arguments `xmin` and `xmax` can be used to subset the data:

```{r, fig.height=4, fig.width=8}
eq_timeline(dataset=eqDataShort, xmin=as.Date("2009-01-01"), xmax=as.Date("2011-01-01"))
```


The function `eq_timeline` is designed to make producing these plots easier, 
but if you want to change the aesthetics arguments or theme, you can use 
the underlying geoms directly, as shown in the following examples.

```{r, fig.height=4, fig.width=8}
ggplot2::ggplot(eqDataShort) +
  geom_timeline(ggplot2::aes(x=DATE, y=Country, fill=TOTAL_DEATHS, size=EQ_PRIMARY))
```

```{r, fig.height=5, fig.width=8}
ggplot2::ggplot(eqDataShort) +
  geom_timeline_label(ggplot2::aes(x=DATE, y=Country, fill=TOTAL_DEATHS, size=EQ_PRIMARY, 
                                   label=LocalLocation))
```

## Interactive Maps

The function `eq_map` produces interactive maps using the JavaScript 
library leaflet (http://leafletjs.com).  Running this function will cause 
a map to open in your default browser. A circle is shown at the epicenter of 
each earthquake with the circle radius being proportional to the magnitude 
(based on the field `EQ_PRIMARY`). The argument `annot_col` should 
be the name of a column (as a string) in the dataset. This column is used 
to generate content for pop-up windows that appear when you click on a 
circle. The following example shows the date in the pop-up window. Click 
on any of the blue circles to see the pop-up window.

```{r, fig.height=5, fig.width=8}
eqDataMexico <- eqDataClean %>%
  dplyr::filter(Country == "Mexico" & YEAR > 2002)
eq_map(dataset=eqDataMexico, annot_col="DATE")
```

The function `eq_create_label` can be used to generate a summary consisting 
of the earthquake's local location (`LocalLocation`), magnitude 
(`EQ_PRIMARY`), and total number of deaths (`TOTAL_DEATHS`). In the 
following example `eq_clean_label` is used to create a new column in the 
data set called `popup_text`, then that column is used in the pop-up window.

```{r, fig.height=5, fig.width=8}
eqDataMexico <- eqDataClean %>%
  dplyr::filter(Country == "Mexico" & YEAR > 2002)
eqDataMexico$popup_text <- eq_create_label(eqDataMexico)
eq_map(dataset=eqDataMexico, annot_col="popup_text")
```
